Implicit neural representation of potential fields

### 2024-06-06 intro.tex
Run spell check

The subsurface continuation of exposed geology can be extrapolated and interpolated using structures mapped in surveys of the Earth's naturally occuring potential fields.
Potential field surveys offer low-cost investigations over a large spatial extent, and are routinely used for geological mapping and interpretation \parencite{nabighian75thAnniversaryHistorical2005}.
For this reason, magnetic field surveys are commonly used by geophysicists for geological mapping and interpretation toward understanding the Earth in three dimensions.
Scattered surveys are commonly interpolated to grid rasters, and interpreted individually or integrated with other datasets for predicting the subsurface for processes such as numerical modelling and inversion.
In these tasks, the spatial resolution of the grid is a critical factor in their ability to define geological detail \parencite{islesGeologicalInterpretationAeromagnetic2013}.
The role of high-resolution details in these geophysical methods is becoming more important to exploration as resources are increasingly sought from below sedimentary cover.
Resolution enhancement for natural images has seen recent success from the use of deep learning \parencite{moserHitchhikerGuideSuperResolution2023}, however, it is not understood if these methods can extend to geophysics data, which differ significantly to the natural images used for training in prior enhancement methods.

@@

\section{Magnetic potential fields}
The Earth has a naturally occuring magnetic field, approximating a dipole oriented toward the top and bottom of the planet.
Several properties of the magnetic field are important to the machine learning methods in this thesis.
First, that a field refers to a set of functions of space and time.
The magnetic field is a vector field, with three orthogonal components.
Neural networks with sufficient parameters are recognised as universal function approximators \parencite{hornikMultilayerFeedforwardNetworks1989}, and should therefore be capable of approximating the functions of the magnetic field.
Second, the curl of the magnetic vector field is a force function known as the magnetic potential field \parencite{kelloggFoundationsPotentialTheory1967}.
The magnitude of the magnetic vector field is the total magnetic intensity (TMI) \parencite{blakelyPotentialTheoryGravity1996}, with the unit nanotesla (nT).

@@

One of the key petrophysical properties influencing the magnetic anomaly is magnetic susceptibility, itself predominantly influenced by magnetite, a mineral commonly present across geological settings \parencite{clarkNotesRockMagnetization1991}.
Rocks may also be intrinsicly magnetised with remnant magnetisation.
Because of these known petrophysical variations in rock magnetisation, surveys of the magnetic field can delineate and differentiate stratigraphic units and infer their composition.

@@

Geoscience data faces a number of challenges distinct from natural image data. These have been outlined in \textcite{karpatneMachineLearningGeosciences2019}, and relevant to this work are:
\begin{itemize}
    \item{} Spatio-temporal structure: Aeromagnetic grids are spatially corellated, neighbouring cells contain similar values arising from the same causative volume.
    \item{} Multi-source multi-resolution data: One location may be covered by multiple aeromagnetic surveys with different resolutions, or the same survey may be presented with different levels of processing.
    \item{} Poor quality of data: Sample data are subject to numerous sources of noise, which combine to become a significant fraction of the gridded data value, or be present as aliased frequencies.
    \item{} Small sample size: While total aeromagnetic coverage in Australia is high, the high-resolution extents are spatially limited and biased by geological domain.
\end{itemize}

@@

\subsection{Deep learning context}
Machine learning (ML) is a broad family of techniques for learning from data.
Artificial neural networks are a subset of techniques in ML, inspired by a model of the biological neuron \parencite{bishopNeuralNetworksPattern1995}.
It has been established that if sufficient conditions are met, neural networks can approximate any function \parencite{hornikMultilayerFeedforwardNetworks1989}.
A mathematical model of a neuron forms the basis of these networks, and the number of neurons within the network and their arrangement are one primary descriptor of a neural network.
Neurons are paremeterised by a multiplicative weight term, an additive bias term, and an activation function which transforms the result.
Each neuron applies these operations to the input, which may either be the input data, or the output of earlier neurons.
The objective of training a neural network is to set the parameters for each neuron in the network, such that the trained network (or model) generalises to novel data.
This is performed by back propagation \parencite{rumelhartLearningRepresentationsBackpropagating1988}, a technique to navigate toward the lowest error solution by calculating the rate of improvement resulting from parameter updates.
A practical introduction to the practice of training machine learning is available in \textcite{stevensDeepLearningPyTorch2020}, while a more formal introduction is provided in \textcite{bishopPatternRecognitionMachine2006}.

>>>>

The subsurface continuation of exposed geology can be extrapolated and interpolated using structures mapped in surveys of the Earth's naturally occurring potential fields.
Potential field surveys offer low-cost investigations over a large spatial extent, and are routinely used for geological mapping and interpretation \parencite{nabighian75thAnniversaryHistorical2005}.
For this reason, magnetic field surveys are commonly used by geophysicists for geological mapping and interpretation toward understanding the Earth in three dimensions.
Scattered surveys are commonly interpolated to grid rasters, and interpreted individually or integrated with other datasets for predicting the subsurface for processes such as numerical modelling and inversion.
In these tasks, the spatial resolution of the grid is a critical factor in their ability to define geological detail \parencite{islesGeologicalInterpretationAeromagnetic2013}.
The role of high-resolution details in these geophysical methods is becoming more important to exploration as resources are increasingly sought from below sedimentary cover.
Resolution enhancement for natural images has seen recent success from the use of deep learning \parencite{moserHitchhikerGuideSuperResolution2023}, however, it is not understood if these methods can extend to geophysics data, which differ significantly to the natural images used for training in prior enhancement methods.

@@

\section{Magnetic potential fields}
The Earth has a naturally occurring magnetic field, approximating a dipole oriented toward the top and bottom of the planet.
Several properties of the magnetic field are important to the machine learning methods in this thesis.
First, that a field refers to a set of functions of space and time.
The magnetic field is a vector field, with three orthogonal components.
Neural networks with sufficient parameters are recognised as universal function approximators \parencite{hornikMultilayerFeedforwardNetworks1989}, and should therefore be capable of approximating the functions of the magnetic field.
Second, the curl of the magnetic vector field is a force function known as the magnetic potential field \parencite{kelloggFoundationsPotentialTheory1967}.
The magnitude of the magnetic vector field is the total magnetic intensity (TMI) \parencite{blakelyPotentialTheoryGravity1996}, with the unit nanotesla (nT).

@@

One of the key petrophysical properties influencing the magnetic anomaly is magnetic susceptibility, itself predominantly influenced by magnetite, a mineral commonly present across geological settings \parencite{clarkNotesRockMagnetization1991}.
Rocks may also be intrinsically magnetised with remnant magnetisation.
Because of these known petrophysical variations in rock magnetisation, surveys of the magnetic field can delineate and differentiate stratigraphic units and infer their composition.

@@

Geoscience data faces a number of challenges distinct from natural image data. These have been outlined in \textcite{karpatneMachineLearningGeosciences2019}, and relevant to this work are:
\begin{itemize}
    \item{} Spatio-temporal structure: Aeromagnetic grids are spatially correlated, neighbouring cells contain similar values arising from the same causative volume.
    \item{} Multi-source multi-resolution data: One location may be covered by multiple aeromagnetic surveys with different resolutions, or the same survey may be presented with different levels of processing.
    \item{} Poor quality of data: Sample data are subject to numerous sources of noise, which combine to become a significant fraction of the gridded data value, or be present as aliased frequencies.
    \item{} Small sample size: While total aeromagnetic coverage in Australia is high, the high-resolution extents are spatially limited and biased by geological domain.
\end{itemize}

@@

\subsection{Deep learning context}
Machine learning (ML) is a broad family of techniques for learning from data.
Artificial neural networks are a subset of techniques in ML, inspired by a model of the biological neuron \parencite{bishopNeuralNetworksPattern1995}.
It has been established that if sufficient conditions are met, neural networks can approximate any function \parencite{hornikMultilayerFeedforwardNetworks1989}.
A mathematical model of a neuron forms the basis of these networks, and the number of neurons within the network and their arrangement are one primary descriptor of a neural network.
Neurons are parameterised by a multiplicative weight term, an additive bias term, and an activation function which transforms the result.
Each neuron applies these operations to the input, which may either be the input data, or the output of earlier neurons.
The objective of training a neural network is to set the parameters for each neuron in the network, such that the trained network (or model) generalises to novel data.
This is performed by back propagation \parencite{rumelhartLearningRepresentationsBackpropagating1988}, a technique to navigate toward the lowest error solution by calculating the rate of improvement resulting from parameter updates.
A practical introduction to the practice of training machine learning is available in \textcite{stevensDeepLearningPyTorch2020}, while a more formal introduction is provided in \textcite{bishopPatternRecognitionMachine2006}.

### 2024-06-06 manuscript.tex
Fix header in bibliography
Just going to ignore header in conclusion reading bibliography.
Run spell check

My first floor colleagues and fellow tech geeks, David, Daniel, Chris, Minh, and Peter --- Your friendship and knowledge has been a wonderful environment to grow in as a geodata scientist.
And all my SES, CET, and friends beyond: Alex, Anne, Ben, France, Jackson, Jess, Jules, Kailah, Lauren, Lauri, Liam, Lilly, Maddie, Malvina, Maria, Nicolas, Ravi, Sean, Sumail, Vickie, and Ze --- Thank you for years of lunches, dinners, beaches, rock climbing, yoga, and all the opportunities to talk rocks that I've been missing.

@@

\smallskip
\medskip
\noindent This research was supported by a Rio Tinto Iron Ore PhD Scholarship, and an Australian Government Research Training Program Scholarship.
\vfill{}

@@

>>>>

My first floor colleagues and fellow tech geeks, David, Daniel, Chris, Minh, and Peter --- Your friendship and knowledge has been a wonderful environment to grow in as a geodata scientist.
All my SES, CET, and friends beyond: Alex, Anne, Ben, France, Jackson, Jess, Joel, Jules, Kailah, Lauren, Lauri, Liam, Lilly, Maddie, Malvina, Maria, Nicolas, Ravi, Sean, Sumail, Vickie, and Ze --- Thank you for years of lunches, dinners, beaches, rock climbing, yoga, and all the opportunities to talk rocks that I've been missing.

@@

\smallskip
\noindent This research was supported by a Rio Tinto Iron Ore PhD Scholarship, and an Australian Government Research Training Program Scholarship.
\vfill{}

@@

\newpagestyle{bib}{
    \headrule
    \sethead[][][]{}{Bibliography}{}
    \setfoot[][][]{}{\thepage}{}
}

\pagestyle{bib}

### 2024-06-06 paper2.tex
Run spell check

Instead of increasing the sampling density, the detail in potential field grids can be enhanced by calculating the potential as if it were sampled at a smaller source-sensor distance, in a process known as downward continuation \parencite{bullardDeterminationMassesNecessary1948}.
At large source-sensor distances, high frequencies are diminished below the noise floor to the point of unobservability, resulting in grids that show only broad scale features, typically representing spatially extensive features deep in the crust.
Downward continuation can be performed with the grid representation of the field, or by modelling causative sources and forward modelling the response at another height \parencite{pilkingtonPotentialFieldContinuation2017}.
In either approach, it is common for noise in the sampled grid product to be significantly increased in the downward continued product, limiting the useful vertical continuation extent of of the method to six times the cell size \parencite{dampneyEquivalentSourceTechnique1969,zuoDownwardContinuationTransformation2020}.
Resolution enhancement through neural network approaches to downward continuation \parencite{liStableDownwardContinuation2023,yeHighprecisionDownwardContinuation2022} is a promising research direction in super-resolution geophysics, but the problem formulation is distinct.
The proposed method attempts to predict missing high frequencies with example based learning from similar geologies, while the prior methods attempt to vertically continue grid data using deep learning methods.

Recently, aeromagnetic resolution enhancement has been performed with deep learning SR \parencite{bavandsavadkoohiHighresolutionAeromagneticMap2023,smithMagneticGridResolution2022}.
The method of \textcite{smithMagneticGridResolution2022} is presented in \Cref{ch:paper1} of this thesis.
These two approaches operate on regional aeromagnetic data at four times scale, using variants of the SRGAN convolutional neural network originally by \textcite{ledigPhotorealisticSingleImage2017}.
The method of \textcite{smithMagneticGridResolution2022} demonstated SR methods are capable of upsampling low-resolution magnetic textures and predicting accurate short wavelengths for these data.
Comparison was made against numerical filters such as the bicubic filter.
However, the low-resolution data were created using a polynomial filter that suppressed high frequencies, rather than a real-world low-resolution geophysical sampling process.
The method of \textcite{bavandsavadkoohiHighresolutionAeromagneticMap2023} uses low- and high-resolution aeromagnetic data from regional surveys in Quebec, Canada, with a realistic low-resolution transform from line spacing and sample altitude differences.
% The dataset contains \qty{267} sample grids from surveys in Quebec, Canada, prepared from surveys with a low-resolution line spacing of \qty{800}{\m}.
The proposed method is distinct from these prior works with its use of extensive and geologically varied synthetic data in coordination with real survey data to improve generalisation for regional aeromagnetics in Australia, and the use of a modern MLP-based SR network.

@@

In \Cref{fig:srdata23}, strong line-perpendicular artefacts and loss of high-frequency detail in LR is lessened in both SR predictions, and fine structural divisions are better delineated.
However, northwest striking features in HR are aliased in LR, and here subsequently incorrectly rotated by 90 degrees in the baseline SR prediction (\Cref{fig:srdata23} d).
The fine-tuned model (\Cref{fig:srdata23} e) resolves the correct strike, and separates the two linear features present.
The enhanced structural similarity is reflected by the higher FSIM score of in the fine-tuned model. 
Similar structural inaccuracies are seen in\Cref{fig:srdata17}, where severe aliasing again fails to be corrected using the baseline model, but is resolved by fune-tuning.

@@

The North-South striking geology in \Cref{fig:srdata18} presents a worst-case scenario for aeromagnetic surveying, where the magnetic intensity varies strongly in the low sampling, line-perpendicular direction.
The low-resolution sampling completely fails to sample all but the largest features with enough lines to represent them in the resulting grid.
Despite this, both models refine the width of these features, approaching the true width represented in HR\@.
The fine structures in HR appear to have limited importance to the SSIM calculation, because the LR and SR grids are reported with a SSIM of 0.9998 or higher.
This suggests that SSIM may be an innapropriate IQA metric in comparing resolution in potential field grids.

>>>>

Instead of increasing the sampling density, the detail in potential field grids can be enhanced by calculating the potential as if it were sampled at a smaller source-sensor distance, in a process known as downward continuation \parencite{bullardDeterminationMassesNecessary1948}.
At large source-sensor distances, high frequencies are diminished below the noise floor to the point of unobservability, resulting in grids that show only broad scale features, typically representing spatially extensive features deep in the crust.
Downward continuation can be performed with the grid representation of the field, or by modelling causative sources and forward modelling the response at another height \parencite{pilkingtonPotentialFieldContinuation2017}.
In either approach, it is common for noise in the sampled grid product to be significantly increased in the downward continued product, limiting the useful vertical continuation extent of the method to six times the cell size \parencite{dampneyEquivalentSourceTechnique1969,zuoDownwardContinuationTransformation2020}.
Resolution enhancement through neural network approaches to downward continuation \parencite{liStableDownwardContinuation2023,yeHighprecisionDownwardContinuation2022} is a promising research direction in super-resolution geophysics, but the problem formulation is distinct.
The proposed method attempts to predict missing high frequencies with example based learning from similar geologies, while the prior methods attempt to vertically continue grid data using deep learning methods.

Recently, aeromagnetic resolution enhancement has been performed with deep learning SR \parencite{bavandsavadkoohiHighresolutionAeromagneticMap2023,smithMagneticGridResolution2022}.
The method of \textcite{smithMagneticGridResolution2022} is presented in \Cref{ch:paper1} of this thesis.
These two approaches operate on regional aeromagnetic data at four times scale, using variants of the SRGAN convolutional neural network originally by \textcite{ledigPhotorealisticSingleImage2017}.
The method of \textcite{smithMagneticGridResolution2022} demonstrated SR methods are capable of upsampling low-resolution magnetic textures and predicting accurate short wavelengths for these data.
Comparison was made against numerical filters such as the bicubic filter.
However, the low-resolution data were created using a polynomial filter that suppressed high frequencies, rather than a real-world low-resolution geophysical sampling process.
The method of \textcite{bavandsavadkoohiHighresolutionAeromagneticMap2023} uses low- and high-resolution aeromagnetic data from regional surveys in Quebec, Canada, with a realistic low-resolution transform from line spacing and sample altitude differences.
% The dataset contains \qty{267} sample grids from surveys in Quebec, Canada, prepared from surveys with a low-resolution line spacing of \qty{800}{\m}.
The proposed method is distinct from these prior works with its use of extensive and geologically varied synthetic data in coordination with real survey data to improve generalisation for regional aeromagnetics in Australia, and the use of a modern MLP-based SR network.

@@

In \Cref{fig:srdata23}, strong line-perpendicular artefacts and loss of high-frequency detail in LR is lessened in both SR predictions, and fine structural divisions are better delineated.
However, northwest striking features in HR are aliased in LR, and here subsequently incorrectly rotated by 90 degrees in the baseline SR prediction (\Cref{fig:srdata23} d).
The fine-tuned model (\Cref{fig:srdata23} e) resolves the correct strike, and separates the two linear features present.
The enhanced structural similarity is reflected by the higher FSIM score of in the fine-tuned model. 
Similar structural inaccuracies are seen in\Cref{fig:srdata17}, where severe aliasing again fails to be corrected using the baseline model, but is resolved by fine-tuning.

@@

The North-South striking geology in \Cref{fig:srdata18} presents a worst-case scenario for aeromagnetic surveying, where the magnetic intensity varies strongly in the low sampling, line-perpendicular direction.
The low-resolution sampling completely fails to sample all but the largest features with enough lines to represent them in the resulting grid.
Despite this, both models refine the width of these features, approaching the true width represented in HR\@.
The fine structures in HR appear to have limited importance to the SSIM calculation, because the LR and SR grids are reported with a SSIM of 0.9998 or higher.
This suggests that SSIM may be an inappropriate IQA metric in comparing resolution in potential field grids.

### 2024-06-06 paper3.tex
Run spell check

\section{Results}
\subsection{Grid prediction from implicit neural representation}
Querying a learned INR using a regularly spaced \(x,\ y\) grid at a specific \(z\) height is a novel method for producing regular grids from geophysical potential field surveys.
The cell size can be selected using a finer or coarser \(x,\ y\) spacing, including variable or anisotropic cell sizes.
Contemporary super-resolution research can leverage this continual representation as a framework for high quality arbitrary scale factor high-frequency signal prediction \parencite[e.g][]{chenLearningContinuousImage2021}.
However, the potential field representation task here does not attempt to predict additional high-frequency information and resolution remains determined by the neural network's capacity to represent high frequencies sampled by the survey line spacing.

>>>>

\section{Results}
\subsection{Grid prediction from implicit neural representation}
Querying a learned INR using a regularly spaced \(x,\ y\) grid at a specific \(z\) height is a novel method for producing regular grids from geophysical potential field surveys.
The cell size can be selected using a finer or coarser \(x,\ y\) spacing, including variable or anisotropic cell sizes.
Contemporary super-resolution research can leverage this continual representation as a framework for high quality arbitrary scale factor high-frequency signal prediction \parencite[e.g.][]{chenLearningContinuousImage2021}.
However, the potential field representation task here does not attempt to predict additional high-frequency information and resolution remains determined by the neural network's capacity to represent high frequencies sampled by the survey line spacing.
